---
title: "Tek_MPP3_MPP4"
author: "Aurelie Lenaerts"
date: "07/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## load libraries

Load the libraries and data required in this project.
```{r libraries, message=FALSE}
library("DESeq2")
library("pheatmap")
library("RColorBrewer")
library("wesanderson")
library("ggplot2")
library("ashr")
library("ggvenn") 
library("viridis") 
library("enrichR")
```
## Chapter 2

Select genes with counts above average expression threshold.
```{r count matrix}
MPP3_counts <- read.delim("../data/Tek_MPP3_rawcounts.tsv")
MPP4_counts <- read.delim("../data/Tek_MPP4_rawcounts.tsv")
genes.filtered <- read.delim("../data/genes.filtered.t2g", header=FALSE)
rawcount_matrix <- merge(MPP3_counts, MPP4_counts)
rownames(rawcount_matrix) <- rawcount_matrix[, "X"]
rawcount_matrix <- rawcount_matrix[, colnames(rawcount_matrix) != "X"]
rawcount_matrix <- as.matrix(rawcount_matrix)
dim(rawcount_matrix)
colnames(rawcount_matrix) <- c("MPP3_KO1", "MPP3_KO2", "MPP3_KO3", "MPP3_KO4", "MPP3_WT1", "MPP3_WT2", "MPP3_WT3", "MPP3_WT4", "MPP4_KO1", "MPP4_KO2", "MPP4_KO3", "MPP4_KO4", "MPP4_WT1", "MPP4_WT2", "MPP4_WT3", "MPP4_WT4")
```

Create samplesheet and rowdata.
```{r samplesheet}
samplesheet <- data.frame(sampleName = colnames(rawcount_matrix))
rownames(samplesheet) <- samplesheet$sampleName
state <- c(rep(0,length(samplesheet$sampleName)))
for (i in 1:length(state))
{
  state[i] <- substr(as.character(samplesheet$sampleName[i]),1,nchar(as.character(samplesheet$sampleName[i]))-1)
}
samplesheet$state <- state
samplesheet
length(unique(samplesheet$state))
stopifnot(all(rownames(samplesheet) == colnames(rawcount_matrix)))
rowdata <- data.frame(ENSEMBL = rownames(rawcount_matrix))
rowdata = merge(rowdata, genes.filtered,
                by.x = "ENSEMBL", by.y = "V2",
                all.x = TRUE, all.y = FALSE, sort = FALSE)
rowdata <- rowdata[!duplicated(rowdata$ENSEMBL),]
```

Create variable of average counts across samples.
```{r avg_cond}
rowdata$max_avg_cond <- NA
for(i in 1:nrow(rowdata)){
  if(i %% 1000 == 0){
    cat(i, " ") # print gene index with a step of 1000
  }
  gene <- rowdata$ENSEMBL[i]
  conditions <- unique(samplesheet$state)
  avg_conds <- unlist(lapply(conditions, function(cond){
    mean(rawcount_matrix[rownames(rawcount_matrix) == gene, grepl(cond, colnames(rawcount_matrix))])
  }))
  rowdata$max_avg_cond[i] <- max(avg_conds)
}
```

Filter for genes with average counts greater than 100 and some data checks.
```{r filter}
avg_cond_thr <- 100
selected_genes <- rowdata$ENSEMBL[rowdata$max_avg_cond > avg_cond_thr]

rawcount_matrix = rawcount_matrix[order(rownames(rawcount_matrix)),]
rowdata = rowdata[order(rowdata$ENSEMBL),]
stopifnot(all(rowdata$ENSEMBL == rownames(rawcount_matrix)))
dim(rawcount_matrix)

rawcount_matrix <- rawcount_matrix[rownames(rawcount_matrix) %in% selected_genes, ]
rowdata <- rowdata[rowdata$ENSEMBL %in% selected_genes, ]
dim(rawcount_matrix)
stopifnot(all(rownames(rawcount_matrix) == rowdata$ENSEMBL))

rownames(rawcount_matrix) <- rowdata$V3

stopifnot(all(rownames(samplesheet) == colnames(rawcount_matrix)))
rownames(rawcount_matrix)[rownames(rawcount_matrix) == ""] 
all(samplesheet$sampleName == colnames(rawcount_matrix))
```
We have 16 conditions and 11831 genes after the threshold.

## Chapter 3

Perform DEseq2 with filtered count data. Relevel the comparisons to contrast MPP3 WT versus KO and MPP4 WT versus KO. Finally, normalised counts are exported and DE genes visualised as volcano plots. 
```{r deseq}
dds_M3_M4 <- DESeqDataSetFromMatrix(rawcount_matrix, samplesheet, design= ~ state, rowData = rowdata)
dds_M3_M4 <- DESeq(dds_M3_M4)

dds_M3_M4$state <- factor(dds_M3_M4$state, levels = c("MPP3_WT","MPP3_KO","MPP4_WT","MPP4_KO"))
dds_M3_M4$state
dds_M3_M4 <- nbinomWaldTest(dds_M3_M4)
resultsNames(dds_M3_M4)

res <- results(dds_M3_M4)
```

```{r normalised counts}
norm_counts <- counts(dds_M3_M4, normalized = TRUE)
norm_counts <- round(norm_counts, 1)
dim(norm_counts)
write.table(norm_counts, file = "../proc_data/MPP3_MPP4_norm_counts.tsv", quote = FALSE, sep = "\t")
```

Visualise MPP3 and MPP4 transcriptomes, using significant DE genes. 
```{r pca, message=FALSE}
vsd <- vst(dds_M3_M4, blind=FALSE)
de <- rownames(res[res$padj<0.1 & !is.na(res$padj),])

PCA <- plotPCA(vsd[de,], intgroup=c("state")) + ylim(-10,10) + xlim(-14,14) + theme_minimal() + 
  theme(axis.line=element_line(colour="black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_color_manual(values=wes_palette("GrandBudapest1"))
print(PCA)
ggsave("../figures/PCA.svg", dpi=300)
```

Comparison of MPP3 and MPP4 transcriptomes in WT conditions.  
```{r WT comparison}
resLFC_M3WTvM4WT <- lfcShrink(dds_M3_M4, contrast = c("state", "MPP3_WT", "MPP4_WT"), type = "ashr")
resOrdered_WT_M3vM4 <- resLFC_M3WTvM4WT[order(resLFC_M3WTvM4WT$log2FoldChange),]
write.table(resOrdered_WT_M3vM4, file = "../proc_data/MPP3WTvMPP4WT_results.tsv", quote = FALSE, sep = "\t")
resSig_WT <- subset(resOrdered_WT_M3vM4, padj < 0.1)
```

```{r WT volcano, message=FALSE}
WT_M3vM4_df <- as.data.frame(resLFC_M3WTvM4WT)
WT_M3vM4_df$padj[is.na(WT_M3vM4_df$padj)] <- 1
WT_M3vM4_df$sigBool = ifelse(WT_M3vM4_df$padj < 0.1, 'S', 'NS')
WT_M3vM4_df$log2FoldChangeADJ <- WT_M3vM4_df$log2FoldChange
WT_M3vM4_df$log2FoldChangeADJ[WT_M3vM4_df$log2FoldChangeADJ > 2] <- 2
WT_M3vM4_df$log2FoldChangeADJ[WT_M3vM4_df$log2FoldChangeADJ < -2] <- -2

WT_MPP3vMPP4_MA <- ggplot(data=WT_M3vM4_df) +
  geom_point_rast(aes(x=log10(baseMean), y=log2FoldChangeADJ, col=sigBool), 
                  alpha=0.5, size=3.75,raster.dpi = 600) +
  theme_minimal() + scale_x_continuous(limits = c(1,5), breaks = c(1, 2,3, 4, 5)) +
  theme(axis.line=element_line(colour="black"), axis.ticks = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("grey","#1f77b4"))
print(WT_MPP3vMPP4_MA)
ggsave("../figures/WT_MPP3vMPP4_MA.svg", dpi=300)
```

Comparison of MPP3 WT versus KO.  
```{r MPP3 WTvKO}
resLFC_M3KOvWT <- lfcShrink(dds_M3_M4, contrast = c("state", "MPP3_KO", "MPP3_WT"), type = "ashr")
resOrdered_M3 <- resLFC_M3KOvWT[order(resLFC_M3KOvWT$log2FoldChange),]
write.table(resOrdered_M3, file = "../proc_data/MPP3_KOvWT_results.tsv", quote = FALSE, sep = "\t")
resSig_M3 <- subset(resOrdered_M3, padj < 0.1)
```

```{r MPP3 volcano, message=FALSE}
M3KOvWT_df <- as.data.frame(resLFC_M3KOvWT)
M3KOvWT_df$padj[is.na(M3KOvWT_df$padj)] <- 1
M3KOvWT_df$sigBool = ifelse(resLFC_M3KOvWT$padj < 0.1, 'S', 'NS')
M3KOvWT_df$log2FoldChangeADJ <- M3KOvWT_df$log2FoldChange
M3KOvWT_df$log2FoldChangeADJ[M3KOvWT_df$log2FoldChangeADJ > 2] <- 2
M3KOvWT_df$log2FoldChangeADJ[M3KOvWT_df$log2FoldChangeADJ < -2] <- -2

MPP3_KOvWT_MA <- ggplot(data=M3KOvWT_df) +
  geom_point_rast(aes(x=log10(baseMean), y=log2FoldChangeADJ, col=sigBool), 
                  alpha=0.5, size=3.75,raster.dpi = 600) +
  theme_minimal() + scale_x_continuous(limits = c(1,6), breaks = c(1, 2,3, 4, 5, 6)) +
  theme(axis.line=element_line(colour="black"), axis.ticks = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("grey","#1f77b4"))
print(MPP3_KOvWT_MA)
ggsave("../figures/MPP3_KOvWT_MA.svg", dpi=300)
```

Comparison of MPP4 WT versus KO.  
```{r MPP4 WTvKO}
resLFC_M4KOvWT <- lfcShrink(dds_M3_M4, contrast = c("state", "MPP4_KO", "MPP4_WT"), type = "ashr")
resOrdered_M4 <- resLFC_M4KOvWT[order(resLFC_M4KOvWT$log2FoldChange),]
write.table(resOrdered_M4, file = "../proc_data/MPP4_KOvWT_results.tsv", quote = FALSE, sep = "\t")
resSig_M4 <- subset(resOrdered_M4, padj < 0.1)
```

```{r MPP4 volcano, message=FALSE}
M4KOvWT_df <- as.data.frame(resLFC_M4KOvWT)
M4KOvWT_df$padj[is.na(M4KOvWT_df$padj)] <- 1
M4KOvWT_df$sigBool = ifelse(M4KOvWT_df$padj < 0.1, 'S', 'NS')
M4KOvWT_df$log2FoldChangeADJ <- M4KOvWT_df$log2FoldChange
M4KOvWT_df$log2FoldChangeADJ[M4KOvWT_df$log2FoldChangeADJ > 2] <- 2
M4KOvWT_df$log2FoldChangeADJ[M4KOvWT_df$log2FoldChangeADJ < -2] <- -2


MPP4_KOvWT_MA <- ggplot(data=M4KOvWT_df) +
  geom_point_rast(aes(x=log10(baseMean), y=log2FoldChangeADJ, col=sigBool), 
                  alpha=0.5, size=3.75,raster.dpi = 600) +
  theme_minimal() + scale_x_continuous(limits = c(1,6), breaks = c(1, 2,3, 4, 5, 6)) +
  theme(axis.line=element_line(colour="black"), axis.ticks = element_line(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_color_manual(values = c("grey","#1f77b4","red"))
print(MPP4_KOvWT_MA)
ggsave("../figures/MPP4_KOvWT_MA.svg", dpi=300)

```

## Chapter 4

z-score heatmap of DRGs.
Create dataframe for MPP3, common and MPP4 DE genes.
```{r DRG dataframe}
M3_DRG <- as.data.frame(resSig_M3)
M4_DRG <- as.data.frame(resSig_M4)
shared_genes <- resSig_M3[rownames(resSig_M3) %in%  rownames(resSig_M4),]
common_DRG <- as.data.frame(shared_genes)
common_DRG$gene <- row.names(common_DRG)
M3_DRG$gene <- row.names(M3_DRG)
M4_DRG$gene <- row.names(M4_DRG)
M3_notshared <- M3_DRG[!(M3_DRG$gene %in% common_DRG$gene), ]
M4_notshared <- M4_DRG[!(M4_DRG$gene %in% common_DRG$gene), ]
M3_spec <- merge(M3_notshared, norm_counts, by = "row.names")
M4_spec <- merge(M4_notshared, norm_counts, by = "row.names")
df_shared <- merge(common_DRG, norm_counts, by = "row.names")
rownames(M3_spec) = M3_spec$Row.names
rownames(M4_spec) = M4_spec$Row.names
rownames(df_shared) <- df_shared$Row.names
```

Calculate z-scores for MPP3 and MPP4 cells separately for each set of DE genes.
```{r M3 DRG zscore}
M3spec_Zscores <- as.data.frame(
  t(apply(M3_spec[,c(8:15)], 1, function(x) (x-mean(x))/sd(x)) ))
M3spec_ZscoresM4 <- as.data.frame(
  t(apply(M3_spec[,c(16:23)], 1, function(x) (x-mean(x))/sd(x)) ))
M3DRG_Zscores <- cbind(M3spec_Zscores, M3spec_ZscoresM4)
M3DRG_Zscores <- M3DRG_Zscores[,c(5,6,7,8,1,2,3,4,13,14,15,16,9,10,11,12)]
```

```{r M4 DRG zscore}
M4spec_Zscores <- as.data.frame(
  t(apply(M4_spec[,c(16:23)], 1, function(x) (x-mean(x))/sd(x)) ))
M4spec_ZscoresM3 <- as.data.frame(
  t(apply(M4_spec[,c(8:15)], 1, function(x) (x-mean(x))/sd(x)) ))
M4DRG_Zscores <- cbind(M4spec_ZscoresM3, M4spec_Zscores)
M4DRG_Zscores <- M4DRG_Zscores[,c(5,6,7,8,1,2,3,4,13,14,15,16,9,10,11,12)]
```

```{r Common DRG zscore}
M4_Zscores <- as.data.frame(t(apply(df_shared[,c(16:23)], 1, function(x) (x-mean(x))/sd(x)) ))
M3_Zscores <- as.data.frame(t(apply(df_shared[,c(8:15)], 1, function(x) (x-mean(x))/sd(x)) ))
shared_Zscores <- cbind(M3_Zscores, M4_Zscores)
shared_Zscores <- shared_Zscores[,c(5,6,7,8,1,2,3,4,13,14,15,16,9,10,11,12)]
```

Extract cluster row order for each set of DE genes. 
```{r M3 DRG roworder}
pal <- wes_palette(name = "Zissou1", 100, type = c("continuous"))
M3_spec_heatmap <- pheatmap(as.matrix(M3DRG_Zscores), cluster_cols=F, cellwidth = 8, cellheight = 2, treeheight_row = 0, show_rownames = FALSE, color = pal, border_color = "black")
M3_spec_heatmap_roworder <- M3DRG_Zscores[M3_spec_heatmap$tree_row$order,]
rev_M3_spec_heatmap_roworder <- apply(M3_spec_heatmap_roworder, 2, rev)
rev_M3_spec_heatmap_roworder <- as.data.frame(rev_M3_spec_heatmap_roworder)
M3_roworder_anno <- data.frame(gene = row.names(rev_M3_spec_heatmap_roworder))
M3_roworder_anno$cluster <- "MPP3"
M3_roworder_anno$gene -> row.names(M3_roworder_anno)
M3_roworder_anno$gene <- NULL
```

```{r M4 DRG roworder}
M4_spec_heatmap <- pheatmap(as.matrix(M4DRG_Zscores), cluster_cols=F, cellwidth = 8, cellheight = 2, treeheight_row = 0, show_rownames = FALSE, color = pal, border_color = "black")
M4_spec_heatmap_roworder <- M4DRG_Zscores[M4_spec_heatmap$tree_row$order,]
rev_M4_spec_heatmap_roworder <- apply(M4_spec_heatmap_roworder, 2, rev)
rev_M4_spec_heatmap_roworder <- as.data.frame(rev_M4_spec_heatmap_roworder)
M4_roworder_anno <- data.frame(gene = row.names(rev_M4_spec_heatmap_roworder))
M4_roworder_anno$cluster <- "MPP4"
M4_roworder_anno$gene -> row.names(M4_roworder_anno)
M4_roworder_anno$gene <- NULL
```

```{r common DRG roworder}
shared_MPP3_4 <- pheatmap(as.matrix(shared_Zscores), cluster_cols=F, cellwidth = 8, cellheight = 2, treeheight_row = 0, legend_breaks = c(-2, 0, 2), show_rownames = FALSE, color = pal)
shared_heatmap_roworder <- shared_Zscores[shared_MPP3_4$tree_row$order,]
common_roworder_anno <- data.frame(gene = row.names(shared_heatmap_roworder))
common_roworder_anno$cluster <- "common"
common_roworder_anno$gene -> row.names(common_roworder_anno)
common_roworder_anno$gene <- NULL
```

Plot heatmap or ordered DRGs.
```{r DRG heatmap}
ordered_DRG_mat <- rbind(rev_M3_spec_heatmap_roworder,shared_heatmap_roworder,rev_M4_spec_heatmap_roworder)
ordered_DRG_anno <- rbind(M3_roworder_anno, common_roworder_anno, M4_roworder_anno)

pheatmap(as.matrix(ordered_DRG_mat), cluster_cols=F, cluster_rows = F, cellwidth = 6, cellheight = 1,
         treeheight_row = 0, annotation_row = ordered_DRG_anno, show_rownames = FALSE, color = pal,
         gaps_row = c(149, 215), gaps_col = c(8), filename = "../figures/DRG_zscore_heatmap.pdf")
```

## Chapter 5

Enrichment analysis of WTvKO and MPP3vMPP4 with enrichR databases. 
```{r enrichrDB}
dbs <- listEnrichrDbs()
dbs <- c("Gene_Perturbations_from_GEO_down", "GO_Biological_Process_2021", "PanglaoDB_Augmented_2021")
```

WT enrichment
```{r WT enrichR}
enriched_WT_M3 <- enrichr(rownames(resSig_WT[resSig_WT$log2FoldChange > 0, ]), dbs)
enriched_WT_M4 <- enrichr(rownames(resSig_WT[resSig_WT$log2FoldChange < 0, ]), dbs)
```
```{r WT GEO down, message=FALSE}
WT_M3_GEO_down = enriched_WT_M3$Gene_Perturbations_from_GEO_down
WT_M3_GEO_down = WT_M3_GEO_down[order(WT_M3_GEO_down$Adjusted.P.value), ]
WT_M3_GEO_down$Combined.Score <- -WT_M3_GEO_down$Combined.Score
WT_M3_GEO_down$Status <- "MPP3"
WT_M4_GEO_down = enriched_WT_M4$Gene_Perturbations_from_GEO_down
WT_M4_GEO_down = WT_M4_GEO_down[order(WT_M4_GEO_down$Adjusted.P.value), ]
WT_M4_GEO_down$Status <- "MPP4"
WT_GEOco <- rbind(WT_M3_GEO_down[c(1:3), ],
                  WT_M4_GEO_down[c(1:3), ])
WT_GEOco$logpadj <- log10(WT_GEOco$Adjusted.P.value)
WT_GEOco$setsize <- sapply(strsplit(WT_GEOco$Overlap, '/'),"[[",1)

M3vM4_GEO_down <- ggplot(WT_GEOco) +
  geom_point(aes(y=reorder(Term, -Combined.Score), x=Combined.Score, color=-logpadj, size = as.numeric(setsize))) +
  scale_color_gradient(low="#9ECAE1", high="#1f77b4") +
  theme_classic()
print(M3vM4_GEO_down)
ggsave("../figures/M3vM4_GEO_down.svg", dpi=300)
```
```{r WT GO_BP, message=FALSE}
WT_M3_GO_BP = enriched_WT_M3$GO_Biological_Process_2021
WT_M3_GO_BP = WT_M3_GO_BP[order(WT_M3_GO_BP$Adjusted.P.value), ]
WT_M3_GO_BP$Combined.Score <- -WT_M3_GO_BP$Combined.Score
WT_M3_GO_BP$Status <- "MPP3"
WT_M4_GO_BP = enriched_WT_M4$GO_Biological_Process_2021
WT_M4_GO_BP = WT_M4_GO_BP[order(WT_M4_GO_BP$Adjusted.P.value), ]
WT_M4_GO_BP$Status <- "MPP4"
WT_GOBPco <- rbind(WT_M3_GO_BP[c(1:3), ],
                   WT_M4_GO_BP[c(1:3), ])
WT_GOBPco$logpadj <- log10(WT_GOBPco$Adjusted.P.value)
WT_GOBPco$setsize <- sapply(strsplit(WT_GOBPco$Overlap, '/'),"[[",1)

M3vM4_GO_BP <- ggplot(WT_GOBPco) +
  geom_point(aes(y=reorder(Term, -Combined.Score), x=Combined.Score, color=-logpadj, size = as.numeric(setsize))) +
  scale_color_gradient(low="#9ECAE1", high="#1f77b4") +
  theme_classic()
print(M3vM4_GO_BP)
ggsave("../figures/M3vM4_GO_BP.svg", dpi=300)
```
```{r WT PanglaoDB, message=FALSE}
WT_M3_celltypes = enriched_WT_M3$PanglaoDB_Augmented_2021
WT_M3_celltypes = WT_M3_celltypes[order(WT_M3_celltypes$Adjusted.P.value), ]
WT_M3_celltypes$Combined.Score <- -WT_M3_celltypes$Combined.Score
WT_M3_celltypes$Status <- "MPP3"
WT_M4_celltypes = enriched_WT_M4$PanglaoDB_Augmented_2021
WT_M4_celltypes = WT_M4_celltypes[order(WT_M4_celltypes$Adjusted.P.value), ]
WT_M4_celltypes$Status <- "MPP4"
WT_Celltypesco <- rbind(WT_M3_celltypes[c(1:3), ],
                        WT_M4_celltypes[c(1:3), ])
WT_Celltypesco$logpadj <- log10(WT_Celltypesco$Adjusted.P.value)
WT_Celltypesco$setsize <- sapply(strsplit(WT_Celltypesco$Overlap, '/'),"[[",1)

M3vM4_Celltypes <- ggplot(WT_Celltypesco) +
  geom_point(aes(y=reorder(Term, -Combined.Score), x=Combined.Score, color=-logpadj, size = as.numeric(setsize))) +
  scale_color_gradient(low="#9ECAE1", high="#1f77b4") +
  theme_classic()
print(M3vM4_Celltypes)
ggsave("../figures/M3vM4_Celltypes.svg", dpi=300)
```
MPP3 WTvKO enrichment
```{r MPP3 enrichR}
enriched_M3KO <- enrichr(rownames(resSig_M3[resSig_M3$log2FoldChange > 0, ]), dbs)
enriched_M3WT <- enrichr(rownames(resSig_M3[resSig_M3$log2FoldChange < 0, ]), dbs)
```
```{r MPP3 PanglaoDB}
M3_Cell_WT = enriched_M3WT$PanglaoDB_Augmented_2021
M3_Cell_WT = M3_Cell_WT[order(M3_Cell_WT$Adjusted.P.value), ]
M3_Cell_WT$Combined.Score <- -M3_Cell_WT$Combined.Score
M3_Cell_WT$Status <- "MPP3_WT"
M3_Cell_KO = enriched_M3KO$PanglaoDB_Augmented_2021
M3_Cell_KO = M3_Cell_KO[order(M3_Cell_KO$Adjusted.P.value), ]
M3_Cell_KO$Status <- "MPP3_KO"
```
```{r MPP3 GO_BP}
M3_BP_WT = enriched_M3WT$GO_Biological_Process_2021
M3_BP_WT = M3_BP_WT[order(M3_BP_WT$Adjusted.P.value), ]
M3_BP_WT$Combined.Score <- -M3_BP_WT$Combined.Score
M3_BP_WT$Status <- "MPP3_WT"
M3_BP_KO = enriched_M3KO$GO_Biological_Process_2021
M3_BP_KO = M3_BP_KO[order(M3_BP_KO$Adjusted.P.value), ]
M3_BP_KO$Status <- "MPP3_KO"
```
```{r MPP3 GEO down}
M3_TF_WT = enriched_M3WT$Gene_Perturbations_from_GEO_down
M3_TF_WT = M3_TF_WT[order(M3_TF_WT$Adjusted.P.value), ]
M3_TF_WT$Combined.Score <- -M3_TF_WT$Combined.Score
M3_TF_WT$Status <- "MPP3_WT"
M3_TF_KO = enriched_M3KO$Gene_Perturbations_from_GEO_down
M3_TF_KO = M3_TF_KO[order(M3_TF_KO$Adjusted.P.value), ]
M3_TF_KO$Status <- "MPP3_KO"
```
```{r M3 enrichR plot, message=FALSE}
M3Co <- rbind(M3_TF_WT[c(1), ],
                 M3_TF_KO[c(1:2), ],
                 M3_BP_WT[c(1), ],
                 M3_BP_KO[c(1:3), ],
                 M3_Cell_WT[c(1:3), ],
                 M3_Cell_KO[c(1:2), ])

M3Co$logpadj <- log10(M3Co$Adjusted.P.value)
M3Co$setsize <- sapply(strsplit(M3Co$Overlap, '/'),"[[",1)

M3_enrichr <- ggplot(M3Co) +
  geom_point(aes(y=reorder(Term, -Combined.Score), x=-logpadj, color=Status, size = as.numeric(setsize))) +
  theme_classic() +
  scale_color_manual(values = c("#e6c19e","#e87b2a"))
print(M3_enrichr)
ggsave("../figures/M3_enrichr.svg", dpi=300)
```
MPP4 WTvKO enrichment
```{r MPP4 enrichR}
enriched_M4KO <- enrichr(rownames(resSig_M4[resSig_M4$log2FoldChange > 0, ]), dbs)
enriched_M4WT <- enrichr(rownames(resSig_M4[resSig_M4$log2FoldChange < 0, ]), dbs)
```
```{r MPP4 PanglaoDB}
M4_Cell_WT = enriched_M4WT$PanglaoDB_Augmented_2021
M4_Cell_WT = M4_Cell_WT[order(M4_Cell_WT$Adjusted.P.value), ]
M4_Cell_WT$Combined.Score <- -M4_Cell_WT$Combined.Score
M4_Cell_WT$Status <- "MPP4_WT"
M4_Cell_KO = enriched_M4KO$PanglaoDB_Augmented_2021
M4_Cell_KO = M4_Cell_KO[order(M4_Cell_KO$Adjusted.P.value), ]
M4_Cell_KO$Status <- "MPP4_KO"
```
```{r MPP4 GO_BP}
M4_BP_WT = enriched_M4WT$GO_Biological_Process_2021
M4_BP_WT = M4_BP_WT[order(M4_BP_WT$Adjusted.P.value), ]
M4_BP_WT$Combined.Score <- -M4_BP_WT$Combined.Score
M4_BP_WT$Status <- "MPP4_WT"
M4_BP_KO = enriched_M4KO$GO_Biological_Process_2021
M4_BP_KO = M4_BP_KO[order(M4_BP_KO$Adjusted.P.value), ]
M4_BP_KO$Status <- "MPP4_KO"
```
```{r MPP4 GEO down}
M4_TF_WT = enriched_M4WT$Gene_Perturbations_from_GEO_down
M4_TF_WT = M4_TF_WT[order(M4_TF_WT$Adjusted.P.value), ]
M4_TF_WT$Combined.Score <- -M4_TF_WT$Combined.Score
M4_TF_WT$Status <- "MPP4_WT"
M4_TF_KO = enriched_M4KO$Gene_Perturbations_from_GEO_down
M4_TF_KO = M4_TF_KO[order(M4_TF_KO$Adjusted.P.value), ]
M4_TF_KO$Status <- "MPP4_KO"
```
```{r M4 enrichR plot, message=FALSE}
M4Co <- rbind(M4_TF_WT[c(1,2), ],
              M4_TF_KO[c(1:3), ],
              M4_Cell_WT[c(1:3), ],
              M4_Cell_KO[c(1,2,4), ])

M4Co$logpadj <- log10(M4Co$Adjusted.P.value)
M4Co$setsize <- sapply(strsplit(M4Co$Overlap, '/'),"[[",1)

M4_enrichr <- ggplot(M4Co) +
  geom_point(aes(y=reorder(Term, -Combined.Score), x=-logpadj, color=Status, size = as.numeric(setsize))) +
  theme_classic() +
  scale_color_manual(values = c("#a8cddf","#1272ba"))
print(M4_enrichr)
ggsave("../figures/M4_enrichr.svg", dpi=300)
```

## Chapter 6 

Cebpa GO heatmap.
Extract DE genes that overlap with Cebpa signature. 
```{r cebpa signature}
GO_TF_trans <- read.csv("../data/GO_TF_trans.csv")
Cebpa_GO_M3 <- resSig_M3[rownames(resSig_M3) %in% GO_TF_trans$Cebpa_ko_mouse_gse61468_creedsid_gene_1476_down,]
Cebpa_GO_M4 <- resSig_M4[rownames(resSig_M4) %in% GO_TF_trans$Cebpa_ko_mouse_gse61468_creedsid_gene_1476_down,]
shared_genes <- resSig_M3[rownames(resSig_M3) %in%  rownames(resSig_M4),]
Cebpa_GO_shared <- shared_genes[rownames(shared_genes) %in% GO_TF_trans$Cebpa_ko_mouse_gse61468_creedsid_gene_1476_down,]
```

Create dataframe for MPP3, common and MPP4 DE genes that overlap with Cebpa signature.

```{r cebpa_dataframe}
M3_cebpa <- data.frame(gene = row.names(Cebpa_GO_M3))
M3_cebpa$factor <- "M3"
M4_cebpa <- data.frame(gene = row.names(Cebpa_GO_M4))
M4_cebpa$factor <- "M4"
common_cebpa <- data.frame(gene = row.names(Cebpa_GO_shared))

cebpa_all <- rbind(M3_cebpa, M4_cebpa)
cebpa_all[(cebpa_all$gene %in% common_cebpa$gene),]$factor <- 'common'
cebpa_all <- cebpa_all[!duplicated(cebpa_all$gene),]
cebpa_all$gene -> row.names(cebpa_all)
cebpa_all$gene <- NULL 

df_cebpa_GO <- merge (cebpa_all, norm_counts, by = "row.names")
target <- c("M3", "common", "M4")
df_cebpa_GO <- df_cebpa_GO %>% arrange(factor(factor, levels = target))
rownames(df_cebpa_GO) = df_cebpa_GO$Row.names

df_cebpa_GO_anno <- df_cebpa_GO[,c(1:2)]
df_cebpa_GO_anno$Row.names <- NULL
```

Calculate z-scores separately for MPP3 and for MPP4 and plot heatmap. 
```{r cebpa_heatmap}
M3_GO_cebpa_Zscores <- as.data.frame(
  t(apply(df_cebpa_GO[,c(3:10)], 1, function(x) (x-mean(x))/sd(x)) ))
M4_GO_cebpa_Zscores <- as.data.frame(
  t(apply(df_cebpa_GO[,c(11:18)], 1, function(x) (x-mean(x))/sd(x)) ))

df_cebpa_GO_zscore <- cbind(M3_GO_cebpa_Zscores, M4_GO_cebpa_Zscores)
df_cebpa_GO_zscore <- df_cebpa_GO_zscore[,c(5,6,7,8,1,2,3,4,13,14,15,16,9,10,11,12)]

pal <- wes_palette(name = "Zissou1", 100, type = c("continuous"))

pheatmap(as.matrix(df_cebpa_GO_zscore), cluster_cols=F, cluster_rows = F, cellwidth = 6, cellheight = 3,
         treeheight_row = 0, annotation_row = df_cebpa_GO_anno, show_rownames = T,
         color = pal, gaps_row = c(36, 53), gaps_col = c(8), filename = "../figures/GO_cebpa_down_MPP3_MPP4.pdf")
```

## Chapter 7

Overlap of heatmaps with cebpa ChIP data.
```{r loda cebpa chip}
cebpa_unionpeaks_finalhits <- 
  read.delim("../data/cebpa_unionpeaks_finalhits.bed", header=FALSE)
cebpa_peaks_anno <- cebpa_unionpeaks_finalhits[, c(1:3,12:13,21:22)]
cebpa_peaks_anno_dedup <- cebpa_peaks_anno[!duplicated(cebpa_peaks_anno$V22), ]
colnames(cebpa_peaks_anno_dedup) <- c("chr","start","end","distTSS","feature","ENS","gene")
```

Which DRGs are cebpa targets?
```{r DRG cebpa targets}
ordered_DRG_mat_cebpa <- ordered_DRG_mat
ordered_DRG_mat_cebpa$gene <- row.names(ordered_DRG_mat_cebpa)
DRG_cebpa<-left_join(x=ordered_DRG_mat_cebpa,y=cebpa_peaks_anno_dedup,by="gene")
DRG_cebpa[is.na(DRG_cebpa)] <- 0
DRG_cebpa$chr[DRG_cebpa$chr > 0] <- 1
identical(DRG_cebpa$gene, ordered_DRG_mat_cebpa$gene)
DRG_cebpa$gene <- factor(DRG_cebpa$gene, levels = as.character(DRG_cebpa$gene))
DRG_cebpa_peak <- ggplot(DRG_cebpa, aes(x = 1, y = gene)) + 
  geom_tile(aes(fill=chr)) +
  scale_y_discrete(limits=rev) + scale_fill_manual(values = c("ivory2", "royalblue3")) +
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../figures/DRG_zscore_heatmap_cebpa_peak.svg", dpi=300)
```

Which DE genes that overlap with cebpa signature genes are cebpa targets?
```{r Cebpa signature cebpa targets}

df_cebpa_GO_zscore_chip <- df_cebpa_GO_zscore
df_cebpa_GO_zscore_chip$gene <- row.names(df_cebpa_GO_zscore_chip)

sign_chip_cebpa<-left_join(x=df_cebpa_GO_zscore_chip,y=cebpa_peaks_anno_dedup,by="gene")
sign_chip_cebpa[is.na(sign_chip_cebpa)] <- 0
sign_chip_cebpa$chr[sign_chip_cebpa$chr > 0] <- 1
identical(sign_chip_cebpa$gene, df_cebpa_GO_zscore_chip$gene)

sign_chip_cebpa$gene <- factor(sign_chip_cebpa$gene, levels = as.character(sign_chip_cebpa$gene))
Cebpa_sign_peak <- ggplot(sign_chip_cebpa, aes(x = 1, y = gene)) + 
  geom_tile(aes(fill=chr)) +
  scale_y_discrete(limits=rev(sign_chip_cebpa$gene)) + scale_fill_manual(values = c("ivory2", "royalblue3")) +
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("../figures/GO_cebpa_down_Cebpa_sign_peak.svg", dpi=300)
```
